---
title: "nextgenCrossWorkbench: a guided workbench for genomic cross prediction and mate allocation"
author: "Mario Morales (front-end) and Sikiru Atanda (backend)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nextgenCrossWorkbench}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**nextgenCrossWorkbench** is a point-and-click [`shiny`](https://shiny.posit.co/)
front-end for **genomic cross prediction** and **mate allocation** in plant and
animal breeding programs. It turns a genomic-selection cross-design pipeline into
a guided, ten-step web application over plain CSV inputs.

The workbench predicts the genetic merit of every possible cross between a set of
genotyped, phenotyped parents, then builds a mating plan that trades off
**genetic gain** against **diversity** (group coancestry / inbreeding). The
breeding-genetics computations are performed by the companion
**`nextgenCrossDesign`** engine (authored by Dr. Sikiru Atanda), which the
workbench drives as a *separate, user-configured R process*. The front-end
installs and launches independently of the backend.

This vignette explains the input files, how to configure and launch the app, and
the analysis workflow — the standard diploid workflow, multi-trait joint
superiority, family-size allocation, the gain–diversity Pareto explorer, robust
(posterior) allocation, and the dosage-aware polyploid and disomic-subgenome
design workflows. The method, metric and breeding-system menus are read from the
backend's capability registry, so they always match the installed engine.

```{r}
library(nextgenCrossWorkbench)
```

## Installation and one-time setup

The front-end and the backend are installed independently:

```{r eval = FALSE}
# 1. the backend engine (performs the computations; a separate R process)
R CMD INSTALL nextgenCrossDesign_0.4.0.tar.gz

# 2. this front-end (pulls its dependencies from CRAN)
install.packages("nextgenCrossWorkbench", dependencies = TRUE)
```

The first time you use the workbench on a machine, create a working folder. This
writes a `config.yml` template and a `runs/` folder for per-run outputs:

```{r eval = FALSE}
init_workbench_dir("~/cross-workbench")
```

Edit `~/cross-workbench/config.yml` so `rscript_path` points at the R
installation that has `nextgenCrossDesign` installed, and (optionally)
`package_library` to the library where it lives. The backend can be a *different*
R installation than the one running the user interface. Every key can also be set
with an environment variable, e.g. `NGCD_RSCRIPT_PATH`.

## Input files

The **standard** (diploid) workflow uses up to four CSV files. Genotype and
phenotype are required; the marker map and trait-direction files are optional.
The package bundles a small demo dataset (10 inbred parents, 12 markers, two
traits) that you can inspect directly:

```{r}
demo_dir <- system.file("app", "data", "demo", package = "nextgenCrossWorkbench")
```

**1. Genotype matrix** — parents in rows, markers in columns, coded 0/1/2 (number
of copies of the alternate allele). The first column holds the parent ID.

```{r}
geno <- read.csv(file.path(demo_dir, "genotype.csv"), check.names = FALSE)
geno[1:5, 1:5]
```

**2. Phenotype table** — one row per parent, one column per trait (matched to the
genotype by ID).

```{r}
pheno <- read.csv(file.path(demo_dir, "phenotype.csv"))
head(pheno)
```

**3. Marker map** (optional) — marker name, chromosome, and position, used to
model recombination when predicting progeny variance.

```{r}
map <- read.csv(file.path(demo_dir, "marker_map.csv"))
head(map)
```

**4. Trait directions** (optional) — the selection direction for each trait, so
that risk traits can be selected *downward*.

```{r}
read.csv(file.path(demo_dir, "trait_direction.csv"))
```

You can upload your own CSVs in the app (comma, semicolon, or tab separated; a
UTF-8 byte-order mark from Excel is handled automatically) or start from this
bundled demo. Every input table is **editable in place** — double-click a cell to
change a value and re-run.

## Launching the workbench

```{r eval = FALSE}
run_workbench("~/cross-workbench")
```

`run_workbench()` opens the app in your browser. (It is wrapped in `\dontrun` in
the help examples because it starts an interactive Shiny session.) You can also
obtain the Shiny app object without launching it — useful for deploying to a
Shiny server:

```{r eval = FALSE}
app <- workbench_app("~/cross-workbench")
```

## The analysis workflow

The app is organized as a ten-step navigation bar; each screen has a collapsible
"how to use this step" guide.

1. **Setup** — confirm the app can reach your R installation and the backend
   engine (green status badges).
2. **Data** — load or upload inputs, map columns (auto-guessed), review
   ID/marker alignment checks, and edit tables. Naming a **target crop** adds a
   crop-suitability note to the results (how well the diploid path is validated for
   that crop archetype); most crops are handled on the standard diploid path.
3. **Objective** — trait-by-trait or index-as-trait prediction; pick which traits
   to include and how to combine them (auto, weighted, economic index, desired
   gain, threshold), and optionally add a **multi-trait joint P(superior progeny)**
   (below).
4. **Scoring** — the trait-value metric (default `var_complex`, the usefulness
   criterion), the breeding system (DH or RIL), recombination and GRM methods,
   and the selection proportion.
5. **QC** — duplicate-parent detection, missingness/MAF filters, optional LD
   pruning, and a residual-heterozygosity audit with a one-click exclusion.
6. **Allocation** — the number of crosses (fixed, or an automatic optimizer that
   sweeps K), the gain-versus-diversity balance, the optimizer, constraints
   on parent use, kinship, quotas, and progeny inbreeding, an optional
   **family-size allocation** (distribute a total progeny budget across the
   plan), and an optional **Pareto frontier explorer** (below).
7. **Advanced** — marker-target steering, lethal-allele handling, cost/logistic
   penalties, budgets, and posterior (MCMC) prediction options.
8. **Output** — whether to write an Excel crossing-plan workbook and PNG figures,
   and the random seed for reproducibility.
9. **Run** — one click assembles a JSON configuration, materializes any edited
   tables, and drives the backend. Errors are surfaced with plain-language hints.
10. **Results** — a KPI row (crosses, mean gain, group coancestry, unique
    parents, maximum parent use, mean progeny inbreeding) above sub-tabs for the
    ranked plan, candidate scores, parent use, **family sizes**, the
    gain–diversity frontier, the **Pareto explorer**, QC audit, input matching,
    marker effects, and method/settings provenance, plus a self-contained
    **interactive HTML report** downloadable as HTML or PDF, and a
    **Diagnostics & tuning** tab (below). When the automatic cross-number
    optimizer runs, a **diminishing-returns chart** (gain vs K with the
    recommended K marked) is shown here directly.

### Diagnostics & tuning

When a result looks off, the **Diagnostics & tuning** tab explains *why* each
procedure produced what it did and *which parameter to change* to steer it. Each
item is graded CHECK (act on it), NOTE (worth knowing), or OK (stable). The most
common example: an automatic cross-number recommendation that lands at the edge
of the swept range means the diminishing-returns elbow was never reached — the
number is capped by your range, not the data — so the diagnostic tells you to
widen "K max" and re-run. Other diagnostics flag a binding pairwise-kinship or
parent-use cap, a trait whose low marker-effect reliability is distorting the
score, disagreement between the robust and point-estimate plans, and blocking QC
issues. The same diagnostics appear in the downloadable report.

### The merit metric: the usefulness criterion

By default a cross is valued by the **usefulness criterion** — the expected mean
of the selected fraction of its progeny, i.e. the progeny mean plus the selection
intensity times the progeny standard deviation. Predicting the progeny *variance*
(which depends on parental complementarity and recombination) is what separates
usefulness from a simple mid-parent value, and is why the marker map matters.
See Lehermeier et al. (2017) and Wolfe et al. (2021) in the references.

### Optimal mate allocation

Given per-cross merit and a relationship matrix, the allocation step selects a
mating plan that maximizes mean merit while constraining group coancestry — an
optimum-contribution-selection philosophy (Meuwissen 1997; Akdemir & Sánchez
2016). Constraints mirror those used in practice: the number of crosses, maximum
and minimum uses per parent, and a maximum pairwise-kinship (culling) threshold.
The **gain–diversity frontier** on the Results screen shows the whole trade-off
curve with your chosen plan marked.

### Automatic cross-number optimization

Rather than fixing the number of crosses, the Allocation screen can **sweep the
number of crosses K** and recommend a value using a diminishing-returns (elbow /
kneedle) rule, an effective-population-size floor, or a coancestry budget. When
the sweep runs, the Results screen shows an interactive **diminishing-returns
chart** — mean gain against K with the recommended K highlighted (plus an
effective-size or coancestry overlay for those rules) — and states the chosen K
and rule, so you can see *why* that number was picked and where extra crosses
stop adding gain.

### Family-size allocation

Once a plan is chosen, the **Family sizes** card on the Allocation screen lets you
enter a **total progeny budget**; the backend distributes it across the selected
crosses in proportion to their merit (score-weighted), with optional minimum and
maximum family sizes. The allocation appears on the **Family sizes** results tab as
a per-cross progeny table and a bar chart — a direct answer to "given N seedlings,
how many should each cross get?".

### Multi-trait joint P(superior progeny)

For multi-trait runs, the **Objective** screen can add a joint superiority column:
the probability that a cross throws progeny clearing your target on *every*
selected trait at once. It uses the estimated cross-trait genetic covariance, so it
captures the joint distribution a per-trait probability misses (a cross can be
strong for one trait yet rarely produce progeny that are simultaneously good for
all). Targets default to each trait's population mean in its selection direction and
can be set per trait. The result is shown as a `p_superior_progeny_mt` column on the
selected-crosses table.

### Pareto frontier explorer

The single gain-versus-diversity dial picks one point automatically. To see the
*whole* trade-off yourself, turn on the **Pareto frontier explorer** on the
Allocation screen and give it an explicit list of diversity-penalty (lambda)
values; the **Pareto explorer** results tab then plots the full frontier (mean gain
vs group coancestry) and lists every point, so you can choose the balance you want
knowingly.

### Robust (posterior) allocation

Because predicted merits carry uncertainty, the workbench can re-optimize on
**posterior quantiles** or **top-N probabilities** instead of point estimates, so
the recommended plan is stable under prediction uncertainty. The robust plan
appears alongside the standard plan on the Results screen.

## Polyploid and clonal design

Switching the **analysis type** to *Polyploid* on the Data screen exposes a
dosage-aware design path for autopolyploids and clonal crops. It takes a dosage
matrix (coded 0..ploidy) and a single-trait phenotype; a marker map and
trait-direction file are not required. The bundled demo is a tetraploid clone
panel:

```{r}
poly_dir <- system.file("app", "data", "demo_poly", package = "nextgenCrossWorkbench")
dosage <- read.csv(file.path(poly_dir, "dosage.csv"), check.names = FALSE)
dosage[1:5, 1:6]
```

Set the ploidy to match your coding (tetraploid data uses 0..4, so Ploidy = 4).
Options include modeling dominance/heterosis (for selecting clones on genotypic
value), double reduction (autopolyploids, ploidy ≥ 4), a ploidy-aware genomic
relationship matrix, and ploidy-aware QC. The resulting plan reports each cross's
predicted mean, variance, dominance-aware usefulness, parent genomic estimated
breeding values, and pairwise kinship.

### Disomic-subgenome (true allopolyploid) design

For **true allopolyploids** — species whose subgenomes are inherited *diploidly*
(each subgenome coded 0..2) — switch the analysis type to *Disomic subgenome*. You
supply a dosage matrix, a single-trait phenotype, and a marker map with a column
that names each marker's subgenome; the workbench splits the data by subgenome and
runs **everything subgenome-aware** — per-subgenome QC, per-subgenome ridge marker
effects, the subgenome genomic-relationship matrix, and subgenome cross scoring —
before a coancestry-aware allocation.

The within-family **usefulness variance is recombination-aware** whenever your
marker map carries a **chromosome column and cM positions** (or bp positions with a
bp-per-cM rate, converted automatically). Each subgenome contributes the exact
doubled-haploid/RIL family variance — the marker-effect contrasts weighted by the
recombination kernel `(1 − 2r) = exp(−2·d/100)` under Haldane, where `d` is the
map distance in cM — and the subgenome contributions are **summed**. That sum is
*exact* under disomic inheritance: homoeologues never pair at meiosis, so there is no
recombination between subgenomes and the cross-subgenome covariance is zero. It is the
*same* recombination-aware calculation the Standard workflow uses, applied per
subgenome. Without chromosome + position columns the app falls back to an
unlinked-loci approximation (reported as `unlinked`) and says so in the run notes. The
relationship matrix follows your chosen **GRM method** (VanRaden or Yang), computed
per subgenome and averaged.

This path is deliberately **not crop-specific**: most polyploid crops (wheat, canola)
are genotyped and analysed as diploid, so use the *Standard* workflow for those and
reserve this path for data you genuinely want modelled as separate diploid subgenomes.
For **autopolyploids** such as potato (tetrasomic, dosages 0..4), use the *Polyploid*
workflow at the appropriate ploidy instead — the disomic-subgenome path is only for
species whose subgenomes each segregate as a diploid.

## Reproducibility and provenance

Set a random **seed** on the Output screen for reproducible optimization.
Every run records its full settings, which are shown on the Results
"Method & settings" tab and embedded in the downloadable report, so a plan can
always be traced back to the exact configuration that produced it.

## References

- Meuwissen, T.H.E., Hayes, B.J., & Goddard, M.E. (2001). Prediction of total
  genetic value using genome-wide dense marker maps. *Genetics* 157(4):1819–1829.
- VanRaden, P.M. (2008). Efficient methods to compute genomic predictions.
  *Journal of Dairy Science* 91(11):4414–4423.
- Meuwissen, T.H.E. (1997). Maximizing the response of selection with a
  predefined rate of inbreeding. *Journal of Animal Science* 75(4):934–940.
- Lehermeier, C., Teyssèdre, S., & Schön, C.-C. (2017). Genetic gain increases by
  applying the usefulness criterion with improved variance prediction in
  selection of crosses. *Genetics* 207(4):1651–1661.
- Akdemir, D. & Sánchez, J.I. (2016). Efficient breeding by genomic mating.
  *Frontiers in Genetics* 7:210.
- Wolfe, M.D., Chan, A.W., Kulakow, P., Rabbi, I., & Jannink, J.-L. (2021).
  Genomic mating in outbred species: predicting cross usefulness with additive
  and total genetic covariance matrices. *Genetics* 219(3):iyab122.
- Amadeu, R.R., et al. (2016). AGHmatrix: R package to construct relationship
  matrices for autotetraploid and diploid species. *The Plant Genome* 9(3).
