---
title: "nextgenCrossWorkbench: a guided workbench for genomic cross prediction and mate allocation"
author: "Mario Morales (front-end) and Sikiru Atanda (backend)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nextgenCrossWorkbench}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**nextgenCrossWorkbench** is a point-and-click [`shiny`](https://shiny.posit.co/)
front-end for **genomic cross prediction** and **mate allocation** in plant and
animal breeding programs. It turns a genomic-selection cross-design pipeline into
a guided, ten-step web application over plain CSV inputs.

The workbench predicts the genetic merit of every possible cross between a set of
genotyped, phenotyped parents, then builds a mating plan that trades off
**genetic gain** against **diversity** (group coancestry / inbreeding). The
breeding-genetics computations are performed by the companion
**`nextgenCrossDesign`** engine (authored by Dr. Sikiru Atanda), which the
workbench drives as a *separate, user-configured R process*. The front-end
installs and launches independently of the backend.

This vignette explains the input files, how to configure and launch the app, and
the analysis workflow — the standard diploid workflow, robust (posterior)
allocation, and the dosage-aware polyploid design workflow.

```{r}
library(nextgenCrossWorkbench)
```

## Installation and one-time setup

The front-end and the backend are installed independently:

```{r eval = FALSE}
# 1. the backend engine (performs the computations; a separate R process)
R CMD INSTALL nextgenCrossDesign_0.4.0.tar.gz

# 2. this front-end (pulls its dependencies from CRAN)
install.packages("nextgenCrossWorkbench", dependencies = TRUE)
```

The first time you use the workbench on a machine, create a working folder. This
writes a `config.yml` template and a `runs/` folder for per-run outputs:

```{r eval = FALSE}
init_workbench_dir("~/cross-workbench")
```

Edit `~/cross-workbench/config.yml` so `rscript_path` points at the R
installation that has `nextgenCrossDesign` installed, and (optionally)
`package_library` to the library where it lives. The backend can be a *different*
R installation than the one running the user interface. Every key can also be set
with an environment variable, e.g. `NGCD_RSCRIPT_PATH`.

## Input files

The **standard** (diploid) workflow uses up to four CSV files. Genotype and
phenotype are required; the marker map and trait-direction files are optional.
The package bundles a small demo dataset (10 inbred parents, 12 markers, two
traits) that you can inspect directly:

```{r}
demo_dir <- system.file("app", "data", "demo", package = "nextgenCrossWorkbench")
```

**1. Genotype matrix** — parents in rows, markers in columns, coded 0/1/2 (number
of copies of the alternate allele). The first column holds the parent ID.

```{r}
geno <- read.csv(file.path(demo_dir, "genotype.csv"), check.names = FALSE)
geno[1:5, 1:5]
```

**2. Phenotype table** — one row per parent, one column per trait (matched to the
genotype by ID).

```{r}
pheno <- read.csv(file.path(demo_dir, "phenotype.csv"))
head(pheno)
```

**3. Marker map** (optional) — marker name, chromosome, and position, used to
model recombination when predicting progeny variance.

```{r}
map <- read.csv(file.path(demo_dir, "marker_map.csv"))
head(map)
```

**4. Trait directions** (optional) — the selection direction for each trait, so
that risk traits can be selected *downward*.

```{r}
read.csv(file.path(demo_dir, "trait_direction.csv"))
```

You can upload your own CSVs in the app (comma, semicolon, or tab separated; a
UTF-8 byte-order mark from Excel is handled automatically) or start from this
bundled demo. Every input table is **editable in place** — double-click a cell to
change a value and re-run.

## Launching the workbench

```{r eval = FALSE}
run_workbench("~/cross-workbench")
```

`run_workbench()` opens the app in your browser. (It is wrapped in `\dontrun` in
the help examples because it starts an interactive Shiny session.) You can also
obtain the Shiny app object without launching it — useful for deploying to a
Shiny server:

```{r eval = FALSE}
app <- workbench_app("~/cross-workbench")
```

## The analysis workflow

The app is organized as a ten-step navigation bar; each screen has a collapsible
"how to use this step" guide.

1. **Setup** — confirm the app can reach your R installation and the backend
   engine (green status badges).
2. **Data** — load or upload inputs, map columns (auto-guessed), review
   ID/marker alignment checks, and edit tables.
3. **Objective** — trait-by-trait or index-as-trait prediction; pick which traits
   to include and how to combine them (auto, weighted, economic index, desired
   gain).
4. **Scoring** — the trait-value metric (default `var_complex`, the usefulness
   criterion), the breeding system (DH or RIL), recombination and GRM methods,
   and the selection proportion.
5. **QC** — duplicate-parent detection, missingness/MAF filters, optional LD
   pruning, and a residual-heterozygosity audit with a one-click exclusion.
6. **Allocation** — the number of crosses (fixed, or an automatic optimizer that
   sweeps K), the gain-versus-diversity balance, the optimizer, and constraints
   on parent use, kinship, quotas, and progeny inbreeding.
7. **Advanced** — marker-target steering, lethal-allele handling, cost/logistic
   penalties, budgets, and posterior (MCMC) prediction options.
8. **Output** — whether to write an Excel crossing-plan workbook and PNG figures,
   and the random seed for reproducibility.
9. **Run** — one click assembles a JSON configuration, materializes any edited
   tables, and drives the backend. Errors are surfaced with plain-language hints.
10. **Results** — a KPI row (crosses, mean gain, group coancestry, unique
    parents, maximum parent use, mean progeny inbreeding) above sub-tabs for the
    ranked plan, candidate scores, parent use, the gain–diversity frontier, QC
    audit, input matching, marker effects, and method/settings provenance, plus a
    self-contained **interactive HTML report** downloadable as HTML or PDF.

### The merit metric: the usefulness criterion

By default a cross is valued by the **usefulness criterion** — the expected mean
of the selected fraction of its progeny, i.e. the progeny mean plus the selection
intensity times the progeny standard deviation. Predicting the progeny *variance*
(which depends on parental complementarity and recombination) is what separates
usefulness from a simple mid-parent value, and is why the marker map matters.
See Lehermeier et al. (2017) and Wolfe et al. (2021) in the references.

### Optimal mate allocation

Given per-cross merit and a relationship matrix, the allocation step selects a
mating plan that maximizes mean merit while constraining group coancestry — an
optimum-contribution-selection philosophy (Meuwissen 1997; Akdemir & Sánchez
2016). Constraints mirror those used in practice: the number of crosses, maximum
and minimum uses per parent, and a maximum pairwise-kinship (culling) threshold.
The **gain–diversity frontier** on the Results screen shows the whole trade-off
curve with your chosen plan marked.

### Automatic cross-number optimization

Rather than fixing the number of crosses, the Allocation screen can **sweep the
number of crosses K** and recommend a value using a diminishing-returns (elbow /
kneedle) rule, an effective-population-size floor, or a coancestry budget. When
the sweep runs, the report adds a diminishing-returns curve and states the chosen
K and rule.

### Robust (posterior) allocation

Because predicted merits carry uncertainty, the workbench can re-optimize on
**posterior quantiles** or **top-N probabilities** instead of point estimates, so
the recommended plan is stable under prediction uncertainty. The robust plan
appears alongside the standard plan on the Results screen.

## Polyploid and clonal design

Switching the **analysis type** to *Polyploid* on the Data screen exposes a
dosage-aware design path for autopolyploids and clonal crops. It takes a dosage
matrix (coded 0..ploidy) and a single-trait phenotype; a marker map and
trait-direction file are not required. The bundled demo is a tetraploid clone
panel:

```{r}
poly_dir <- system.file("app", "data", "demo_poly", package = "nextgenCrossWorkbench")
dosage <- read.csv(file.path(poly_dir, "dosage.csv"), check.names = FALSE)
dosage[1:5, 1:6]
```

Set the ploidy to match your coding (tetraploid data uses 0..4, so Ploidy = 4).
Options include modeling dominance/heterosis (for selecting clones on genotypic
value), double reduction (autopolyploids, ploidy ≥ 4), a ploidy-aware genomic
relationship matrix, and ploidy-aware QC. The resulting plan reports each cross's
predicted mean, variance, dominance-aware usefulness, parent genomic estimated
breeding values, and pairwise kinship.

## Reproducibility and provenance

Set a random **seed** on the Output screen for reproducible optimization.
Every run records its full settings, which are shown on the Results
"Method & settings" tab and embedded in the downloadable report, so a plan can
always be traced back to the exact configuration that produced it.

## References

- Meuwissen, T.H.E., Hayes, B.J., & Goddard, M.E. (2001). Prediction of total
  genetic value using genome-wide dense marker maps. *Genetics* 157(4):1819–1829.
- VanRaden, P.M. (2008). Efficient methods to compute genomic predictions.
  *Journal of Dairy Science* 91(11):4414–4423.
- Meuwissen, T.H.E. (1997). Maximizing the response of selection with a
  predefined rate of inbreeding. *Journal of Animal Science* 75(4):934–940.
- Lehermeier, C., Teyssèdre, S., & Schön, C.-C. (2017). Genetic gain increases by
  applying the usefulness criterion with improved variance prediction in
  selection of crosses. *Genetics* 207(4):1651–1661.
- Akdemir, D. & Sánchez, J.I. (2016). Efficient breeding by genomic mating.
  *Frontiers in Genetics* 7:210.
- Wolfe, M.D., Chan, A.W., Kulakow, P., Rabbi, I., & Jannink, J.-L. (2021).
  Genomic mating in outbred species: predicting cross usefulness with additive
  and total genetic covariance matrices. *Genetics* 219(3):iyab122.
- Amadeu, R.R., et al. (2016). AGHmatrix: R package to construct relationship
  matrices for autotetraploid and diploid species. *The Plant Genome* 9(3).
