---
title: "nextgenCrossWorkbench: a guided workbench for genomic cross prediction and mate allocation"
author: "Mario Morales (front-end) and Sikiru Atanda (backend)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nextgenCrossWorkbench}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**nextgenCrossWorkbench** is a point-and-click [`shiny`](https://shiny.posit.co/)
front-end for **genomic cross prediction** and **mate allocation** in plant and
animal breeding programs. It turns a genomic-selection cross-design pipeline into
a guided web application over plain CSV inputs, organized as four workflow stages —
**Data · Configure · Run · Results**.

The workbench predicts the genetic merit of every possible cross between a set of
genotyped, phenotyped parents, then builds a mating plan that trades off
**genetic gain** against **diversity** (group coancestry / inbreeding). The
breeding-genetics computations are performed by the companion
**`nextgenCrossDesign`** engine (authored by Dr. Sikiru Atanda), which the
workbench drives as a *separate, user-configured R process*. The front-end
installs and launches independently of the backend.

This vignette explains the input files, how to configure and launch the app, and
the analysis workflow — the standard diploid workflow, the three-way selection
objective (single trait, a computed multi-trait index, or your own index column),
multi-trait joint superiority, per-trait check-line vetoes, the cross-priority
**portfolio & risk** view, family-size allocation, the gain–diversity Pareto
explorer, robust (posterior) allocation, and the dosage-aware polyploid and
disomic-subgenome design workflows. The method, metric, parent-type and
breeding-system menus are read from the backend's capability registry, so they
always match the installed engine; any capability the backend marks
*experimental* or *guarded* is automatically hidden from the menus.

```{r}
library(nextgenCrossWorkbench)
```

## Installation and one-time setup

The front-end and the backend are installed independently:

```{r eval = FALSE}
# 1. the backend engine (performs the computations; a separate R process)
R CMD INSTALL nextgenCrossDesign_0.17.2.tar.gz

# 2. this front-end (pulls its dependencies from CRAN)
install.packages("nextgenCrossWorkbench", dependencies = TRUE)
```

The first time you use the workbench on a machine, create a working folder. This
writes a `config.yml` template and a `runs/` folder for per-run outputs:

```{r eval = FALSE}
init_workbench_dir("~/cross-workbench")
```

Edit `~/cross-workbench/config.yml` so `rscript_path` points at the R
installation that has `nextgenCrossDesign` installed, and (optionally)
`package_library` to the library where it lives. The backend can be a *different*
R installation than the one running the user interface. Every key can also be set
with an environment variable, e.g. `NGCD_RSCRIPT_PATH`.

## Input files

The **standard** (diploid) workflow uses up to four CSV files. Genotype and
phenotype are required; the marker map and trait-direction files are optional.
The package bundles a small demo dataset (10 inbred parents, 12 markers, two
traits) that you can inspect directly:

```{r}
demo_dir <- system.file("app", "data", "demo", package = "nextgenCrossWorkbench")
```

**1. Genotype matrix** — parents in rows, markers in columns, coded 0/1/2 (number
of copies of the alternate allele). The first column holds the parent ID.

```{r}
geno <- read.csv(file.path(demo_dir, "genotype.csv"), check.names = FALSE)
geno[1:5, 1:5]
```

**2. Phenotype table** — one row per parent, one column per trait (matched to the
genotype by ID).

```{r}
pheno <- read.csv(file.path(demo_dir, "phenotype.csv"))
head(pheno)
```

**3. Marker map** (optional) — marker name, chromosome, and position, used to
model recombination when predicting progeny variance.

```{r}
map <- read.csv(file.path(demo_dir, "marker_map.csv"))
head(map)
```

**4. Trait directions** (optional) — the selection direction for each trait, so
that risk traits can be selected *downward*.

```{r}
read.csv(file.path(demo_dir, "trait_direction.csv"))
```

You can upload your own CSVs in the app (comma, semicolon, or tab separated; a
UTF-8 byte-order mark from Excel is handled automatically) or start from this
bundled demo. Every input table is **editable in place** — double-click a cell to
change a value and re-run.

## Launching the workbench

```{r eval = FALSE}
run_workbench("~/cross-workbench")
```

`run_workbench()` opens the app in your browser. (It is wrapped in `\dontrun` in
the help examples because it starts an interactive Shiny session.) You can also
obtain the Shiny app object without launching it — useful for deploying to a
Shiny server:

```{r eval = FALSE}
app <- workbench_app("~/cross-workbench")
```

## The analysis workflow

The app is organized as **four workflow stages** in the top navigation bar —
**Data · Configure · Run · Results** — that follow the order you actually work in:
load and check your data, configure the design, run it, then explore the plan.
Every screen has a collapsible "how to use this section" guide (collapsed by
default; a single **Show guided tour** toggle opens them all).

**Data** — load or upload inputs, map columns (auto-guessed), review ID/marker
alignment checks, and edit tables in place. Naming a **target crop** adds a
crop-suitability note to the results (how well the diploid path is validated for
that crop archetype); most crops are handled on the standard diploid path. A
**Data quality** sub-tab holds duplicate-parent detection, missingness/MAF
filters, optional LD pruning, and a residual-heterozygosity audit with a
one-click exclusion.

**Configure** groups every design choice into five sections, in pipeline order:

1. **Selection objective** — the breeder question "what does *good* mean?" as a
   three-way choice: a **single trait**, **multiple traits (build a selection
   index)**, or **use my selection-index column** (a pre-computed index already in
   your phenotype file). Single- and multiple-trait modes both run full
   trait-by-trait prediction; for multiple traits you pick how they combine
   (automatic, relative weights, economic weights, desired gains) and can add a
   **multi-trait joint P(superior progeny)** column (below). Economic-weight and
   desired-gain *values* come from your trait-direction file.
2. **Prediction & scoring** — the effect & variance model (recombination and GRM
   methods, marker-effect reliability floor, optional training-set augmentation
   and posterior engine), the **cross-scoring metric** (breeder-framed choices —
   *Mid-parent mean*, *Family variance*, *Reliable family variance*, *Usefulness*
   (the default), *Parent distance*) with its joint-superiority option, the
   breeding system (DH or RIL) with the selection proportion, and the **Parent
   type** (*Inbred* / *DH* / *RIL*). Parent type describes the parents you cross,
   not the progeny: *Inbred* and *DH* expect fully fixed lines and block
   heterozygous parents as a data error, while *RIL* accepts the residual
   heterozygosity recombinant inbred lines retain after finite selfing (if Data
   flagged non-inbred parents, either drop them on the Data screen or set Parent
   type = RIL here).
3. **Cross filters & genetic constraints** — candidate-level screens applied
   *before* allocation: the **per-trait check-line veto** (below), lethal-allele
   guarding, and marker-target steering.
4. **Mate allocation** — how the mating plan is built, in three panels: *Plan
   size & constraints* (number of crosses — fixed or an automatic optimizer that
   sweeps K — parent-use caps, kinship cap, quotas, family-size budget),
   *Gain–diversity & relatedness* (the gain-versus-diversity dial, the unified
   **mate-relatedness** control, and the Pareto explorer), and *Engine &
   advanced* (the optimizer, allocation method, robust/posterior allocation).
5. **Export options** — whether to write an Excel crossing-plan workbook and PNG
   figures, and the random seed for reproducibility.

**Run** — the standard workflow runs as four explicit, **manually gated** steps
rather than one monolithic run: **Run QC → Fit effects & score → Build selection
index → Allocate & rank**. Each step's button unlocks only once its upstream step
has completed, so you walk the pipeline deliberately; nothing runs automatically.
QC is a real gate — it blocks the design only on *blockers* (a duplicate-genotype
or other hard issue disables the downstream steps), while warnings pass through.
Each stage's result is cached, so a downstream stage never recomputes upstream
work, and changing a setting marks only that stage and the ones after it as
needing a re-run; staged execution is byte-identical to the old one-shot run. Each
activity screen carries a collapsible **Figure** tag that reveals that step's
chart in place (putative duplicates after QC, trait-model reliability after Fit &
score, the index distribution after Build index, the gain–diversity frontier and
parent use after Allocate & rank). Every step assembles a JSON configuration,
materializes any edited tables, and drives the backend out-of-process; errors and
backend warnings are surfaced with plain-language hints. (Automatic cross-number
selection and the exported workbook/figures still use the full one-shot run.)

**Results** — a KPI row (crosses, mean gain, group coancestry, unique parents,
maximum parent use, mean progeny inbreeding) above sub-tabs for the ranked plan,
candidate scores, parent use, **family sizes**, the **Portfolio & risk** view
(below), the gain–diversity frontier, the **Pareto explorer**, QC audit, input
matching, marker effects, and method/settings provenance, plus a self-contained
**interactive HTML report** downloadable as HTML or PDF, and a **Diagnostics &
tuning** tab (below). When the automatic cross-number optimizer runs, a
**diminishing-returns chart** (gain vs K with the recommended K marked) is shown
here directly.

### Diagnostics & tuning

When a result looks off, the **Diagnostics & tuning** tab explains *why* each
procedure produced what it did and *which parameter to change* to steer it. Each
item is graded CHECK (act on it), NOTE (worth knowing), or OK (stable). The most
common example: an automatic cross-number recommendation that lands at the edge
of the swept range means the diminishing-returns elbow was never reached — the
number is capped by your range, not the data — so the diagnostic tells you to
widen "K max" and re-run. Other diagnostics flag a binding pairwise-kinship or
parent-use cap, a trait whose low marker-effect reliability is distorting the
score, disagreement between the robust and point-estimate plans, how many
candidate crosses a **trait-check veto** flagged or removed, the mix of
**portfolio profiles** and estimation-risk bins in the plan, and blocking QC
issues. The same diagnostics appear in the downloadable report.

### The merit metric: the usefulness criterion

By default a cross is valued by the **usefulness criterion** — the expected mean
of the selected fraction of its progeny, i.e. the progeny mean plus the selection
intensity times the progeny standard deviation. Predicting the progeny *variance*
(which depends on parental complementarity and recombination) is what separates
usefulness from a simple mid-parent value, and is why the marker map matters.
See Lehermeier et al. (2017) and Wolfe et al. (2021) in the references.

### Optimal mate allocation

Given per-cross merit and a relationship matrix, the allocation step selects a
mating plan that maximizes mean merit while constraining group coancestry — an
optimum-contribution-selection philosophy (Meuwissen 1997; Akdemir & Sánchez
2016). Constraints mirror those used in practice: the number of crosses, maximum
and minimum uses per parent, and a maximum pairwise-kinship (culling) threshold.
The **gain–diversity frontier** on the Results screen shows the whole trade-off
curve with your chosen plan marked.

### Unified mate-relatedness control

Relatedness between the two parents of a cross can be steered with a single
control in *Mate allocation → Gain–diversity & relatedness*, rather than by
juggling separate penalties. **Avoid inbreeding** discourages crossing close
relatives (useful for keeping progeny heterozygosity up); **favor
complementarity** does the opposite where a program wants to concentrate
favorable alleles; **off** leaves per-cross relatedness unpenalized (group
coancestry is still constrained by the gain–diversity dial). A single weight sets
the strength. The control is deliberately one dial: the backend rejects a
configuration that also sets the older separate mating / progeny-inbreeding
penalties, so you cannot silently stack two relatedness terms.

### Automatic cross-number optimization

Rather than fixing the number of crosses, the Allocation screen can **sweep the
number of crosses K** and recommend a value using a diminishing-returns (elbow /
kneedle) rule, an effective-population-size floor, or a coancestry budget. When
the sweep runs, the Results screen shows an interactive **diminishing-returns
chart** — mean gain against K with the recommended K highlighted (plus an
effective-size or coancestry overlay for those rules) — and states the chosen K
and rule, so you can see *why* that number was picked and where extra crosses
stop adding gain.

### Family-size allocation

Once a plan is chosen, the **Family sizes** card on the Allocation screen lets you
enter a **total progeny budget**; the backend distributes it across the selected
crosses in proportion to their merit (score-weighted), with optional minimum and
maximum family sizes. The allocation appears on the **Family sizes** results tab as
a per-cross progeny table and a bar chart — a direct answer to "given N seedlings,
how many should each cross get?".

### Multi-trait joint P(superior progeny)

For multi-trait runs, the **Objective** screen can add a joint superiority column:
the probability that a cross throws progeny clearing your target on *every*
selected trait at once. It uses the estimated cross-trait genetic covariance, so it
captures the joint distribution a per-trait probability misses (a cross can be
strong for one trait yet rarely produce progeny that are simultaneously good for
all). Targets default to each trait's population mean in its selection direction and
can be set per trait. The result is shown as a `p_superior_progeny_mt` column on the
selected-crosses table.

### Per-trait check-line veto

Breeders optimize a single objective but still judge crosses **trait by trait**.
The **Cross filters & genetic constraints** section lets you set, for any trait, a
**check line** — a reference parent (or check variety) — and flag every cross
whose mid-parent value for that trait falls on the *worse* side of the check. The
comparison is made on a consistent basis you choose (**GEBV** or **phenotype**),
and the worse side follows the trait's selection direction (increase → reject
below the check; decrease → reject above). By default violators are only
**flagged** (so you see them on the plan with per-trait check columns); ticking
**exclude threshold violators** drops them from the candidate pool *before*
allocation, exactly like the lethal-allele guard. The objective itself is never
touched — this is a candidate-eligibility filter that works with any multi-trait
method — and it is available in the single- and multiple-trait objective modes
(not when you supply your own pre-computed index column). A run note reports how
many crosses were flagged or removed.

### Pareto frontier explorer

The single gain-versus-diversity dial picks one point automatically. To see the
*whole* trade-off yourself, turn on the **Pareto frontier explorer** on the
Allocation screen and give it an explicit list of diversity-penalty (lambda)
values; the **Pareto explorer** results tab then plots the full frontier (mean gain
vs group coancestry) and lists every point, so you can choose the balance you want
knowingly.

### Robust (posterior) allocation

Because predicted merits carry uncertainty, the workbench can re-optimize on
**posterior quantiles** or **top-N probabilities** instead of point estimates, so
the recommended plan is stable under prediction uncertainty. The robust plan
appears alongside the standard plan on the Results screen.

### Portfolio & risk

Two crosses with the same score can be very different bets. The **Portfolio &
risk** results tab (for single-trait runs) decomposes each cross into two purely
genetic axes: its **level** (the mid-parent breeding value — how good it is on
average) and its **upside** (the within-family genetic standard deviation, √VPM —
how much a lucky segregant could exceed the mid-parent), colored by an
**estimation-risk** bin (how confident the prediction is, from the mid-parent
prediction-error variance). Iso-usefulness contours mark equal-merit lines, so you
can read each cross's *profile* at a glance — a high-level/low-spread
**workhorse**, a high-upside **breakthrough** bet, a **long shot**, or a cross to
**deprioritize** — instead of a single opaque ranking. This is a decision-support
view layered on top of the priority tiers; it does not change how crosses are
scored or allocated. Run notes summarize the plan's portfolio mix and estimation
risk, and the same summary appears in the downloadable report.

## Polyploid and clonal design

Switching the **analysis type** to *Polyploid* on the Data screen exposes a
dosage-aware design path for autopolyploids and clonal crops. It takes a dosage
matrix (coded 0..ploidy) and a single-trait phenotype; a marker map and
trait-direction file are not required. The bundled demo is a tetraploid clone
panel:

```{r}
poly_dir <- system.file("app", "data", "demo_poly", package = "nextgenCrossWorkbench")
dosage <- read.csv(file.path(poly_dir, "dosage.csv"), check.names = FALSE)
dosage[1:5, 1:6]
```

Set the ploidy to match your coding (tetraploid data uses 0..4, so Ploidy = 4).
Options include modeling dominance/heterosis (for selecting clones on genotypic
value), double reduction (autopolyploids, ploidy ≥ 4), a ploidy-aware genomic
relationship matrix, and ploidy-aware QC. The resulting plan reports each cross's
predicted mean, variance, dominance-aware usefulness, parent genomic estimated
breeding values, and pairwise kinship.

### Disomic-subgenome (true allopolyploid) design

For **true allopolyploids** — species whose subgenomes are inherited *diploidly*
(each subgenome coded 0..2) — switch the analysis type to *Disomic subgenome*. You
supply a dosage matrix, a single-trait phenotype, and a marker map with a column
that names each marker's subgenome; the workbench splits the data by subgenome and
runs **everything subgenome-aware** — per-subgenome QC, per-subgenome ridge marker
effects, the subgenome genomic-relationship matrix, and subgenome cross scoring —
before a coancestry-aware allocation.

The within-family **usefulness variance is recombination-aware** whenever your
marker map carries a **chromosome column and cM positions** (or bp positions with a
bp-per-cM rate, converted automatically). Each subgenome contributes the exact
doubled-haploid/RIL family variance — the marker-effect contrasts weighted by the
recombination kernel `(1 − 2r) = exp(−2·d/100)` under Haldane, where `d` is the
map distance in cM — and the subgenome contributions are **summed**. That sum is
*exact* under disomic inheritance: homoeologues never pair at meiosis, so there is no
recombination between subgenomes and the cross-subgenome covariance is zero. It is the
*same* recombination-aware calculation the Standard workflow uses, applied per
subgenome. Without chromosome + position columns the app falls back to an
unlinked-loci approximation (reported as `unlinked`) and says so in the run notes. The
relationship matrix follows your chosen **GRM method** (VanRaden or Yang), computed
per subgenome and averaged.

This path is deliberately **not crop-specific**: most polyploid crops (wheat, canola)
are genotyped and analysed as diploid, so use the *Standard* workflow for those and
reserve this path for data you genuinely want modelled as separate diploid subgenomes.
For **autopolyploids** such as potato (tetrasomic, dosages 0..4), use the *Polyploid*
workflow at the appropriate ploidy instead — the disomic-subgenome path is only for
species whose subgenomes each segregate as a diploid.

## Reproducibility and provenance

Set a random **seed** on the Output screen for reproducible optimization.
Every run records its full settings, which are shown on the Results
"Method & settings" tab and embedded in the downloadable report, so a plan can
always be traced back to the exact configuration that produced it.

## References

- Meuwissen, T.H.E., Hayes, B.J., & Goddard, M.E. (2001). Prediction of total
  genetic value using genome-wide dense marker maps. *Genetics* 157(4):1819–1829.
- VanRaden, P.M. (2008). Efficient methods to compute genomic predictions.
  *Journal of Dairy Science* 91(11):4414–4423.
- Meuwissen, T.H.E. (1997). Maximizing the response of selection with a
  predefined rate of inbreeding. *Journal of Animal Science* 75(4):934–940.
- Lehermeier, C., Teyssèdre, S., & Schön, C.-C. (2017). Genetic gain increases by
  applying the usefulness criterion with improved variance prediction in
  selection of crosses. *Genetics* 207(4):1651–1661.
- Akdemir, D. & Sánchez, J.I. (2016). Efficient breeding by genomic mating.
  *Frontiers in Genetics* 7:210.
- Wolfe, M.D., Chan, A.W., Kulakow, P., Rabbi, I., & Jannink, J.-L. (2021).
  Genomic mating in outbred species: predicting cross usefulness with additive
  and total genetic covariance matrices. *Genetics* 219(3):iyab122.
- Amadeu, R.R., et al. (2016). AGHmatrix: R package to construct relationship
  matrices for autotetraploid and diploid species. *The Plant Genome* 9(3).
